### 2025-07-18 paper/quarto_aaai/paper.tex
final adjustments to text

Specifically, \textbf{we propose Counterfactual Training (CT)}: a novel
training regime that leverages counterfactual explanations on-the-fly to
ensure that differentiable models learn plausible and actionable
explanations for the underlying data, while at the same time also being
more robust to adversarial examples (AE). Figure~\ref{fig-poc}
illustrates the outcome of CT compared to conventionally trained models:
in panel (a), faithful and valid counterfactuals end up near the
decision boundary forming a clearly distinguishable cluster in the
target class (orange); in panel (b), CT is applied to the same
underlying linear classifier architecture resulting in much more
plausible counterfactuals; in panel (c), the classifier is again trained
conventionally and we have introduced a mutability constraint on the
\emph{age} feature at test time---counterfactuals are valid but the
classifier is roughly equally sensitive to both features; by contrast,
the decision boundary in panel (c) has titled, making the model trained
with CT relatively less sensitive to the immutable \emph{age} feature.
To achieve these outcomes, CT draws inspiration from the literature on
contrastive and robust learning: we contrast faithful (\emph{mature})
CEs with ground-truth data while protecting immutable features, and
capitalize on methodological links between CE and AE by penalizing the
model's adversarial loss on interim (\emph{nascent}) counterfactuals. To
the best of our knowledge, CT represents the first venture in this
direction with promising empirical and theoretical results.

@@

Our focus is the improvement in explainability (Def.
\ref{def-explainability}). Thus, we primarily look at the plausibility
and cost of faithfully generated counterfactuals at test time. Other
metrics, such as validity and redundancy, are reported in the
supplementary appendix. To measure the cost, we follow the standard
proxy of distances (\(\ell_1\)-norm) between factuals and
counterfactuals. For plausibility, we assess how similar CEs are to the
observed samples in the target domain,
\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the metric used by
\citet{altmeyer2024faithful},
\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation} and introduce a novel divergence metric,
\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
counterfactuals and \(\text{MMD}(\cdot)\) is the unbiased estimate of
the squared population maximum mean discrepancy, proposed by
\citet{gretton2012kernel}. The metric in Equation~\ref{eq-impl-div} is
equal to zero if and only if the two distributions are exactly the same,
\(\mathbf{X}^\prime=\mathbf{X}^+\).

For predictive performance, we use standard metrics, such as robust
accuracy estimated on adversarially perturbed data using FGSM
\citep{goodfellow2014explaining}.

@@

Our main quantitative results for \emph{MLP} models are summarised in
Table~\ref{tbl-main}, which presents average outcomes along with
bootstrapped two standard errors. The following example motivates CT and
illustrates how to read Table~\ref{tbl-main}.

@@

\caption{\label{tbl-main}Key counterfactual evaluation metrics and
bootstrapped standard errors for all datasets. \textbf{Plausibility}
(columns 1-2): percentage reduction in implausibility for \(\text{IP}\)
and \(\text{IP}^*\), respectively; \textbf{Actionability} (column 3):
percentage reduction in costs with protected features. Outcomes are
aggregated across bootstrap samples and varying degrees of the energy
penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time.
Asterisks (\(^*\)) indicate that bootstrapped 99\%-confidence interval
of differences in mean outcomes does \emph{not} include zero.}

@@

We also find that CT can reduce sensitivity to immutable, protected
features and thus lead to less costly counterfactual outcomes as shown
in Figure~\ref{fig-poc}. In column three of Table~\ref{tbl-main}, we
impose mutability constraints on selected features and compute the
reduction in average costs of CEs associated with CT compared to the
baseline: for synthetic datasets, we always protect the first feature;
for all real-world tabular datasets we could identify and protect an
\emph{age} variable; for \emph{MNIST}, we protect the five upper and
lower pixel rows of the full image. Reductions in costs are
overwhelmingly positive and significant of up to nearly 60\% for
\emph{GMSC}. While the estimated cost reductions for \emph{Adult} and
\emph{MNIST} are not significant, Figure~\ref{fig-mnist} (columns 6-10)
demonstrates that CT does have the expected effect: sensitivity to
protected features as per integrated gradients is drastically reduced;
details of this experiment are reported in the supplementary appendix.
In the case of \emph{Cred}, average costs increase, likely because any
potential benefits from protecting the \emph{age} are outweighed by the
increase in costs required for greater plausibility.

>>>>

Specifically, we propose \textbf{counterfactual training (CT)}: a novel
training regime that leverages counterfactual explanations on-the-fly to
ensure that differentiable models learn plausible and actionable
explanations for the underlying data, while at the same time also being
more robust to adversarial examples (AE). Figure~\ref{fig-poc}
illustrates the outcome of CT compared to conventionally trained models:
in panel (a), faithful and valid counterfactuals end up near the
decision boundary forming a clearly distinguishable cluster in the
target class (orange); in panel (b), CT is applied to the same
underlying linear classifier architecture resulting in much more
plausible counterfactuals; in panel (c), the classifier is again trained
conventionally and we have introduced a mutability constraint on the
\emph{age} feature at test time---counterfactuals are valid but the
classifier is roughly equally sensitive to both features; by contrast,
the decision boundary in panel (c) has titled, making the model trained
with CT relatively less sensitive to the immutable \emph{age} feature.
To achieve these outcomes, CT draws inspiration from the literature on
contrastive and robust learning: we contrast faithful (\emph{mature})
CEs with ground-truth data while protecting immutable features, and
capitalize on methodological links between CE and AE by penalizing the
model's adversarial loss on interim (\emph{nascent}) counterfactuals. To
the best of our knowledge, CT represents the first venture in this
direction with promising empirical and theoretical results.

@@

Our focus is the improvement in explainability (Def.
\ref{def-explainability}). Thus, we primarily look at the plausibility
and cost of faithfully generated counterfactuals at test time. Other
metrics, such as validity and redundancy, are reported in the
supplementary appendix. To measure the cost, we follow the standard
proxy of distances (\(\ell_1\)-norm) between factuals and
counterfactuals. For plausibility, we assess how similar CEs are to the
observed samples in the target domain,
\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the metric used by
\citet{altmeyer2024faithful},
\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation} and introduce a novel divergence-based adaption,
\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
counterfactuals and \(\text{MMD}(\cdot)\) is the unbiased estimate of
the squared population maximum mean discrepancy, proposed by
\citet{gretton2012kernel}. The metric in Equation~\ref{eq-impl-div} is
equal to zero if and only if the two distributions are exactly the same,
\(\mathbf{X}^\prime=\mathbf{X}^+\).

To assess outcomes with respect to actionability for non-linear models,
we look at the average costs of valid counterfactuals in terms of their
distances from factual starting points. While this an imperfect proxy of
sensitivity, we hypothesize that CT can reduce these costs by teaching
models to seek plausibility with respect to mutable features, much like
we observe in Figure~\ref{fig-poc} (panel (d) vs panel (c)). We
supplement this analysis with qualitative findings for integrated
gradients \citep{sundararajan2017ig}. Finally, for predictive
performance, we use standard metrics, such as robust accuracy estimated
on adversarially perturbed data using FGSM
\citep{goodfellow2014explaining}.

@@

Our main results for plausibility and actionability for \emph{MLP}
models are summarised in Table~\ref{tbl-main}, which presents
counterfactual outcomes grouped by dataset along with standard errors
averaged across bootstrap samples.

The first two columns (\(\text{IP}\) and \(\text{IP}^*\)) show the
percentage reduction in implausibility for our two metrics when using CT
on top of the weak baseline. Focusing on the first row for \emph{LS} for
a moment, the observed positive values indicate that faithful
counterfactuals are around 30-55\% more plausible for models trained
with CT, in line with our observations in Figure~\ref{fig-poc} (panel
(b) vs.~(a)). Asterisks indicate that reductions are statistically
significant: bootstrapped 99\%-confidence intervals of differences in
mean outcomes do \emph{not} include zero.

The third column shows results for the scenario when mutability
constraints are imposed on selected features. Once again we are
comparing CT to the baseline, so reductions in the positive direction
imply that valid counterfactuals are cheaper when using CT with feature
protection. Relating this back to Figure~\ref{fig-poc}, the third column
represents the reduction in distances travelled by counterfactuals in
panel (d) compared to (c). In the following, we summarize the remaining
results.

@@

\caption{\label{tbl-main}Key evaluation metrics for valid counterfactual
along with bootstrapped standard errors for all datasets.
\textbf{Plausibility} (columns 1-2): percentage reduction in
implausibility for \(\text{IP}\) and \(\text{IP}^*\), respectively;
\textbf{Actionability} (column 3): percentage reduction in costs with
protected features. Outcomes are aggregated across bootstrap samples and
varying degrees of the energy penalty \(\lambda_{\text{egy}}\) used for
\emph{ECCo} at test time. Asterisks (\(^*\)) indicate that bootstrapped
99\%-confidence interval of differences in mean outcomes does \emph{not}
include zero.}

@@

We also find that CT can improve actionability in the presence of
immutable features (column 3 in Table~\ref{tbl-main}): for synthetic
datasets, we always protect the first feature; for all real-world
tabular datasets we could identify and protect an \emph{age} variable;
for \emph{MNIST}, we protect the five upper and lower pixel rows of the
full image. Reductions in costs are overwhelmingly positive and
significant of up to nearly 60\% for \emph{GMSC}. While the estimated
cost reductions for \emph{Adult} and \emph{MNIST} are not significant,
Figure~\ref{fig-mnist} (columns 6-10) demonstrates that in the latter
case, CT does have the expected effect: sensitivity to protected
features as per integrated gradients is drastically reduced; details of
this experiment are reported in the supplementary appendix. In the case
of \emph{Cred}, average costs increase, likely because any potential
benefits from protecting the \emph{age} are outweighed by the increase
in costs required for greater plausibility.

### 2025-07-18 paper/sections/experiments.qmd
final adjustments to text

Our focus is the improvement in explainability (Def. \ref{def-explainability}). Thus, we primarily look at the plausibility and cost of faithfully generated counterfactuals at test time. Other metrics, such as validity and redundancy, are reported in the supplementary appendix. To measure the cost, we follow the standard proxy of distances ($\ell_1$-norm) between factuals and counterfactuals. For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the metric used by @altmeyer2024faithful,
$$
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
$$ {#eq-impl-dist}
and introduce a novel divergence metric,
$$
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
$$ {#eq-impl-div}
where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is the unbiased estimate of the squared population maximum mean discrepancy, proposed by @gretton2012kernel. The metric in @eq-impl-div is equal to zero if and only if the two distributions are exactly the same, $\mathbf{X}^\prime=\mathbf{X}^+$.

For predictive performance, we use standard metrics, such as robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].

@@

Our main quantitative results for *MLP* models are summarised in @tbl-main, which presents average outcomes along with bootstrapped two standard errors. The following example motivates CT and illustrates how to read @tbl-main.

@@

Key counterfactual evaluation metrics and bootstrapped standard errors for all datasets. **Plausibility** (columns 1-2): percentage reduction in implausibility for $\text{IP}$ and $\text{IP}^*$, respectively; **Actionability** (column 3): percentage reduction in costs with protected features. Outcomes are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time. Asterisks ($^*$) indicate that bootstrapped 99%-confidence interval of differences in mean outcomes does *not* include zero.

@@

We also find that CT can reduce sensitivity to immutable, protected features and thus lead to less costly counterfactual outcomes as shown in @fig-poc. In column three of @tbl-main, we impose mutability constraints on selected features and compute the reduction in average costs of CEs associated with CT compared to the baseline: for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five upper and lower pixel rows of the full image. Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*. While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist (columns 6-10) demonstrates that CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced; details of this experiment are reported in the supplementary appendix. In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.

>>>>

Our focus is the improvement in explainability (Def. \ref{def-explainability}). Thus, we primarily look at the plausibility and cost of faithfully generated counterfactuals at test time. Other metrics, such as validity and redundancy, are reported in the supplementary appendix. To measure the cost, we follow the standard proxy of distances ($\ell_1$-norm) between factuals and counterfactuals. For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the metric used by @altmeyer2024faithful,
$$
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
$$ {#eq-impl-dist}
and introduce a novel divergence-based adaption,
$$
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
$$ {#eq-impl-div}
where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is the unbiased estimate of the squared population maximum mean discrepancy, proposed by @gretton2012kernel. The metric in @eq-impl-div is equal to zero if and only if the two distributions are exactly the same, $\mathbf{X}^\prime=\mathbf{X}^+$.

To assess outcomes with respect to actionability for non-linear models, we look at the average costs of valid counterfactuals in terms of their distances from factual starting points. While this an imperfect proxy of sensitivity, we hypothesize that CT can reduce these costs by teaching models to seek plausibility with respect to mutable features, much like we observe in @fig-poc (panel (d) vs panel (c)). We supplement this analysis with qualitative findings for integrated gradients [@sundararajan2017ig]. Finally, for predictive performance, we use standard metrics, such as robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].

@@

Our main results for plausibility and actionability for *MLP* models are summarised in @tbl-main, which presents counterfactual outcomes grouped by dataset along with standard errors averaged across bootstrap samples.

The first two columns ($\text{IP}$ and $\text{IP}^*$) show the percentage reduction in implausibility for our two metrics when using CT on top of the weak baseline. Focusing on the first row for *LS* for a moment, the observed positive values indicate that faithful counterfactuals are around 30-55% more plausible for models trained with CT, in line with our observations in @fig-poc (panel (b) vs. (a)). Asterisks indicate that reductions are statistically significant: bootstrapped 99%-confidence intervals of differences in mean outcomes do *not* include zero.

The third column shows results for the scenario when mutability constraints are imposed on selected features. Once again we are comparing CT to the baseline, so reductions in the positive direction imply that valid counterfactuals are cheaper when using CT with feature protection. Relating this back to @fig-poc, the third column represents the reduction in distances travelled by counterfactuals in panel (d) compared to (c). In the following, we summarize the remaining results.

@@

Key evaluation metrics for valid counterfactual along with bootstrapped standard errors for all datasets. **Plausibility** (columns 1-2): percentage reduction in implausibility for $\text{IP}$ and $\text{IP}^*$, respectively; **Actionability** (column 3): percentage reduction in costs with protected features. Outcomes are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time. Asterisks ($^*$) indicate that bootstrapped 99%-confidence interval of differences in mean outcomes does *not* include zero.

@@

We also find that CT can improve actionability in the presence of immutable features (column 3 in @tbl-main): for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five upper and lower pixel rows of the full image. Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*. While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist (columns 6-10) demonstrates that in the latter case, CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced; details of this experiment are reported in the supplementary appendix. In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.

### 2025-07-18 paper/sections/intro.qmd
final adjustments to text

Specifically, **we propose Counterfactual Training (CT)**: a novel training regime that leverages counterfactual explanations on-the-fly to ensure that differentiable models learn plausible and actionable explanations for the underlying data, while at the same time also being more robust to adversarial examples (AE). @fig-poc illustrates the outcome of CT compared to conventionally trained models: in panel (a), faithful and valid counterfactuals end up near the decision boundary forming a clearly distinguishable cluster in the target class (orange); in panel (b), CT is applied to the same underlying linear classifier architecture resulting in much more plausible counterfactuals; in panel (c), the classifier is again trained conventionally and we have introduced a mutability constraint on the *age* feature at test time---counterfactuals are valid but the classifier is roughly equally sensitive to both features; by contrast, the decision boundary in panel (c) has titled, making the model trained with CT relatively less sensitive to the immutable *age* feature. To achieve these outcomes, CT draws inspiration from the literature on contrastive and robust learning: we contrast faithful (*mature*) CEs with ground-truth data while protecting immutable features, and capitalize on methodological links between CE and AE by penalizing the model's adversarial loss on interim (*nascent*) counterfactuals. To the best of our knowledge, CT represents the first venture in this direction with promising empirical and theoretical results.

>>>>

Specifically, we propose **counterfactual training (CT)**: a novel training regime that leverages counterfactual explanations on-the-fly to ensure that differentiable models learn plausible and actionable explanations for the underlying data, while at the same time also being more robust to adversarial examples (AE). @fig-poc illustrates the outcome of CT compared to conventionally trained models: in panel (a), faithful and valid counterfactuals end up near the decision boundary forming a clearly distinguishable cluster in the target class (orange); in panel (b), CT is applied to the same underlying linear classifier architecture resulting in much more plausible counterfactuals; in panel (c), the classifier is again trained conventionally and we have introduced a mutability constraint on the *age* feature at test time---counterfactuals are valid but the classifier is roughly equally sensitive to both features; by contrast, the decision boundary in panel (c) has titled, making the model trained with CT relatively less sensitive to the immutable *age* feature. To achieve these outcomes, CT draws inspiration from the literature on contrastive and robust learning: we contrast faithful (*mature*) CEs with ground-truth data while protecting immutable features, and capitalize on methodological links between CE and AE by penalizing the model's adversarial loss on interim (*nascent*) counterfactuals. To the best of our knowledge, CT represents the first venture in this direction with promising empirical and theoretical results.
